---
title: "03.1.2_rothc_FYM-FB"
author: Tobias Ruhnau
date: 06.11.2024
output: html_document
editor_options: 
  chunk_output_type: console
---

Experiment: PK+ME-FB (ME = FYM)

Script to run rothc with data from literature research and database: EJPSoil_BUFACAP_template.xlsx

This script only treats experiment: Farako-Ba, Burkina Faso (SSA)

Created by T. Ruhnau (22.10.2024, edited 06.11.2024 and 11.11.2024)

#-- Preperation ---------------------------------------------------------------#
Read packages and input data -> data comes from 02_climate_data.R and 01_auto_database_FB.R
```{r libraries}
library(SoilR)
library(readr)
library(dplyr)
library(tidyverse)
```

The information of vegetation periods has to be taken directly from the literature or NDVI values, if not available in LTFEs

Split climate data up for historical spin-up (1970-1990) and prediction model (1991-2024).
If there is climate data available in the corresponding literature, it should be used.
Normally, no evaporation data available. Hence, it is derived from ERA5 via GEE.
(The split-up assignment is random and follows no peer-reviewed basis < SOURCE?!)

According to Peña Cabriales et al. (2006) on pg. 59 it says: 
"one rainy season per year, starting in May-June and ending in October." 
> Filter is set accordingly.

Always run in the beginning
```{r climate monthly means}
# Monthly means from climate_data_FB.R
mm1 <- read.csv2("output/farako_ba/climate/mm_1970_1990.csv", header = TRUE) %>% dplyr::select(-1)
mm2 <- read.csv2("output/farako_ba/climate/mm_1991_2024.csv", header = TRUE) %>% dplyr::select(-1)

# Filter Month and Experiment for Spinup section (1970 - 1990)
mm1_spinup <- mm1 %>%
  # filter(Month >= 5 & Month <= 10) %>% 
  filter(Experiment == "Farako-Ba") # Choose the experiment

# # Filter Month and Experiment for Section1 Rain (1991 - 2024)
# mm2_sec1 <- mm2 %>% # This will be used for Model Section
#   filter(Month >= 5 & Month <= 10) %>%
#   filter(Experiment == "Farako-Ba") # Choose the experiment
# 
# # Filter Month and Experiment for Section2 Dry (1991 - 2024)
# mm2_sec2 <- mm2 %>% # This will be used for Model Section
#   filter(Month %in% c(11:12, 1:4)) %>% # to filter months through year
#   filter(Experiment == "Farako-Ba") # Choose the experiment

# Filter only for experiment (not dividing into sections)
mm2_pred <- mm2 %>% 
  filter(Experiment == "Farako-Ba")

rm(mm1, mm2)
```

This next two chunks only need to run once to extract relevant soil input parameters such as 
C input, initial SOC stock (year of first measurement), soil depth and clay content.
```{r read soil data}
# Specify the directory where your CSV files are located
csv_dir <- "output/farako_ba/" # change dir to different LTFEs

# Get a list of all CSV files in the directory
csv_files <- list.files(path = csv_dir, pattern = "*.csv", full.names = TRUE)

# Read all CSV files into a list of dataframes
df_list <- map(csv_files, ~read_csv2(.) %>% dplyr::select(-1)) # -1 to get rid of first column

# Name the dataframes based on the file names (without extension)
names(df_list) <- tools::file_path_sans_ext(basename(csv_files))

# Unpack list to global environment
list2env(df_list, .GlobalEnv)

rm(df_list)
```

```{r soil prep C-FB}
### This chunk selects all parameters for the treatment C-FB (control site in Farako-Ba)
## For further information see literature in database

# Calculate soil.thick
fb_datasoil <- fb_datasoil %>% 
  mutate(soil.thick = Depth_to_cm - Depth_from_cm) # Indicates soil depth

# Get initial SOC stock 
fb_SOC <- fb_datasoil %>% 
  filter(tID == "PK+ME-FB") %>% # Change to respective treatment
  dplyr::select(eID, tID, Sample_year, SOC_stock_Mg.ha) # Use first year as initial SOC stock

# Get clay content
fb_clay <- fb_soiltype %>% 
  filter(Bottom_depth_cm == 20) # Use value for clay model

# Get carbon residue input
fb_cinput <- fb_datacrop %>% 
  dplyr::select(eID, tID, Residues_above_Mg.ha, DC_Comment) %>% 
  filter(tID == "PK+ME-FB") %>% # Change to respective treatment
  mutate(MeanResidues = mean(Residues_above_Mg.ha)) # Use mean for Cinputs in Model
```

#-- Spin-up section -----------------------------------------------------------#

```{r climate spin-up}
## Climate data for Farako-ba, Burkina Faso for 1970-1990 (ERA5)

# Create a new order of months to fit Souther Hemisphere
new_order <- c(7:12, 1:6) # this is essential to avoid negative respiration in fw_Model

# Mean monthly temperature for 1970-1990
Temp_Spinup <- mm1_spinup %>% 
  filter(Variable == "Temp") %>% 
  dplyr::select(Month, MeanValue) %>% 
  rename(Temp = MeanValue)

# Reindex the dataframe to start with July (7)
Temp_Spinup <- Temp_Spinup[order(match(Temp_Spinup$Month, new_order)), ]

# Mean monthly precipitation for 1970-1990
Prec_Spinup <- mm1_spinup %>% 
  filter(Variable == "Prec") %>% 
  dplyr::select(Month, MeanValue) %>% 
  rename(Prec = MeanValue)

# Reindex the dataframe to start with July (7)
Prec_Spinup <- Prec_Spinup[order(match(Prec_Spinup$Month, new_order)), ]

# Mean monthly evaporation for 1970-1990
Evap_Spinup <- mm1_spinup %>% 
  filter(Variable == "Evapo") %>% 
  dplyr::select(Month, MeanValue) %>% 
  rename(Evap = MeanValue)# Reindex the dataframe to start with July (7)

Evap_Spinup <- Evap_Spinup[order(match(Evap_Spinup$Month, new_order)), ]

# Try to convert to open-pan evaporation by multiplying with 0.75
Evap_Spinup$Evap <- 0.75*Evap_Spinup$Evap # suggested by Coleman et al. 1996 (doi: 10.1007/978-3-642-61094-3_17)
```

```{r soil spin-up}
# Prepare soil data
soil.thick_Spinup=20 # Same as model
SOC_Spinup=0         # Change to fit initial SOC stock of model ---> 12 for PK+ME-FB
clay_Spinup=7        # Same as model
Cinputs_Spinup=2.3   # 2.3 (11.11.24)
# Former land use = Natural grassland
DPM.RPM = 0.67       # DPM/RPM --> DR = arable: 1.44, grasslands: 0.67, forest: 0.25 (Gottschalk et al. 2012)
FYM = 0              # Input factor for Farmyard manure by time

# Create vector of time steps to run spin-up
years_Spinup=seq(from=0, to=500, by=1/12)
```

```{r prepare spin-up}
# Calculate effects of climate on decomposition
fT_Spinup=fT.RothC(Temp_Spinup[,2]) # Temperature effects per month 
fW_Spinup=fW.RothC(P=(Prec_Spinup[,2]), E=(Evap_Spinup[,2]),
            S.Thick = soil.thick_Spinup, pClay = clay_Spinup,
            pE = 1.0, bare = FALSE)$b # Rainy season = vegetation covering the soil (bare = F)

# # Replace negative values in fW with 0
# fW_Spinup[fW_Spinup < 0] <- 0 # high Evapo values leads to error in model, still not sure if this is the right way to handle it

# >>Add comment<<
xi.frame_Spinup=data.frame(years_Spinup,rep(fT_Spinup*fW_Spinup,length.out=length(years_Spinup)))

# Estimating IOM using Falloon method
FallIOM=0.049*SOC_Spinup^(1.139) 
```

```{r run spin-up}
# Run the model Section1
Model_Spinup=RothCModel(t=years_Spinup,C0=c(DPM=0, RPM=0, BIO=0, HUM=0, IOM=FallIOM), 
                  In=Cinputs_Spinup, clay=clay_Spinup, xi=xi.frame_Spinup,
                  DR = DPM.RPM, FYM = FYM)

# Calculates SOC stocks for each pool
Ct_Spinup=getC(Model_Spinup) 

# Final pool size spin-up
poolSize_Spinup=as.numeric(tail(Ct_Spinup,1)) # Uses the last values to show final pool size
names(poolSize_Spinup)<-c("DPM", "RPM", "BIO", "HUM", "IOM") # Is that correct?

# Print pool size for spin-up section
poolSize_Spinup

# Sum the carbon stocks for total spin up
total_C_Spinup <- sum(poolSize_Spinup)
rows_C_Spinup <- rowSums(Ct_Spinup)

# Combine total carbon stocks into a data frame
total_C_df <- data.frame(
    Model = c("Spinup"),
    Total_C = c(sum(poolSize_Spinup))
)

# Print total sum of C after spin-up
print(total_C_df)
```

#-- Prediction section --------------------------------------------------------#

```{r climate model}
### Climate data for Farako-ba, Burkina Faso 
## IMPORTANT: "Thus, in the Southern Hemisphere, the weather data file should start in July when the soil is wet, so that July will appear as January in the output." (coleman and Jenkinson 1996)
# Use data from peer-reviewed papers, if existing.

# Mean monthly temperature for 1991-2024 (ERA5)
Temp_Model <- mm2_pred %>%
  filter(Variable == "Temp") %>%
  dplyr::select(Month, MeanValue) %>%
  rename(Temp = MeanValue)

# Reindex the dataframe to start with July (7)
Temp_Model <- Temp_Model[order(match(Temp_Model$Month, new_order)), ]

# Mean monthly precipitation for 1991-2024 (ERA5)
Prec_Model <- mm2_pred %>%
  filter(Variable == "Prec") %>%
  dplyr::select(Month, MeanValue) %>%
  rename(Prec = MeanValue)

# Reindex the dataframe to start with July (7)
Prec_Model <- Prec_Model[order(match(Prec_Model$Month, new_order)), ]

# Change certain values of Precipitation to actual measurements from literature
Prec_Model1 <- Prec_Model %>%
  mutate(Prec = case_when(
    Month == 5 ~ 109,
    Month == 6 ~ 118,
    Month == 7 ~ 197,
    Month == 8 ~ 255,
    Month == 9 ~ 189,
    Month == 10 ~ 68,
    TRUE ~ Prec
  ))
# Reassing for further processing
Prec_Model <- Prec_Model1

# Mean monthly evaporation for 1991-2024 (ERA5)
Evap_Model <- mm2_pred %>%
  filter(Variable == "Evapo") %>%
  dplyr::select(Month, MeanValue) %>%
  rename(Evap = MeanValue)

# Reindex the dataframe to start with July (7)
Evap_Model <- Evap_Model[order(match(Evap_Model$Month, new_order)), ]

# Try to convert to open-pan evaporation by multiplying with 0.75
Evap_Model$Evap <- 0.75*Evap_Model$Evap # suggested by Coleman et al. 1996 (doi: 10.1007/978-3-642-61094-3_17)
```

```{r soil model}
### Soil data
soil.thick_Model = 20    # Soil thickness (organic layer topsoil), in cm
SOC_Model = 12           # Value of Spin-up Section should fit this initial SOC stock
clay_Model = 7           # Percent clay
# Cinputs_Model = 2.510889 # 2.510889 for C-FB # Annual C inputs to soil in Mg/ha/yr
DPM.RPM = 1.44           # DPM/RPM --> DR = arable: 1.44, grasslands: 0.67, forest: 0.25 (Gottschalk et al. 2012)
# FYM = 0                # FYM allocation: DPM: 49%, RPM: 49%, HUM: 2% (Weihermueller et al. 2013)

# Create vector of time steps to run model
years_Model = seq(1/12, 20, by=1/12)

# Create df for C inputs, it replaces Cinputs_Model
df_cinputs <- data.frame(
  month = years_Model, # PK should be recognized in residue ammendment
  values = c(0, 0, 0, 0, 4.95, 4.95, 4.95, 4.95, 4.95, 0, 0, 0) # Each value represents one month in the year
) # Mean residues for treatment, crops average = SSS, GSC, SF, G, C in CMS, M in CMS, S in CMS (database)

# Create df for FYM inputs, it replaces FYM
df_fym <- data.frame(
  month = years_Model,
  values = c(0, 0, 0, 0, 0.24, 0.24, 0.24, 0.24, 0.24, 0, 0, 0) # Each value represents one month in the year
) # Check database > 3*0.4 = 1.2 Mg/ha/y FYM application (0.4 represents C content of FYM < SOURCE?!)
# 1.2/5 = 0.24 for each month where soil was covered with crops
```

```{r prepare model}
# Calculate effects of climate on decomposition
fT_Model=fT.RothC(Temp_Model[,2]) # Temperature effects per month 
fW_Model=fW.RothC(P=(Prec_Model[,2]), E=(Evap_Model[,2]), # Rain and evaporation effects
            S.Thick = soil.thick_Model, pClay = clay_Model,
            pE = 1.0, bare = F)$b # bare = T or F is covered in df_cinputs above
# 
# # Replace negative values in fW with 0
# fW_Model[fW_Model < 0] <- 0 # high Evap values leads to error in model

xi.frame_Model=data.frame(years_Model, rep(fT_Model*fW_Model, length.out=length(years_Model)))

# Estimating IOM using Falloon method
FallIOM=0.049*SOC_Model^(1.139)
```

```{r run model}
# Run the model
Soil_Model <- RothCModel(
  t=years_Model, # Use input from soil model # Numeric values #
  ks = c(k.DPM = 10, k.RPM = 0.3, k.BIO = 0.66, k.HUM = 0.02, k.IOM = 0), # Use default values
  C0 = poolSize_Spinup,# Use input from spin up section # Pools #
  In = df_cinputs, # Use input from soil model # Mg/ha #
  FYM = df_fym, # Use input from soil model # Mg/ha #
  DR = DPM.RPM, # Use input from soil model # Ratio #
  clay = clay_Model, # Use input from soil model # % #
  xi = xi.frame_Model, # Use input from prepare model
  # solver = deSolve.lsoda.wrapper, # For non-linear calculations only
  pass = FALSE
)

# Computes pools as function of time
Ct_Soil=getC(Soil_Model)

poolSize_Model=as.numeric(tail(Ct_Soil,1))
names(poolSize_Model)<-c("DPM", "RPM", "BIO", "HUM", "IOM")
poolSize_Model
```

```{r pool model}
# Sum of SOC stocks per row
rows_C_Model <- rowSums(Ct_Soil)

# Sum of SOC stocks
total_C_Model <- sum(poolSize_Model)
total_C_Model 

# Create a data frame for the final pool sizes
final_pools_df <- data.frame(
    Pool = c("DPM", "RPM", "BIO", "HUM", "IOM"),
    Model = poolSize_Model,
    Spin_up = poolSize_Spinup
)

# Combine total carbon stocks into a data frame with spin-up
total_C_df <- data.frame(
    Model = c("Model", "Spin_up"),
    Total_C = c(sum(poolSize_Model), sum(poolSize_Spinup)),
    Years = c(20, 500) #
)

# Print results
print(total_C_df)

# Determine the maximum y-axis limit to fit both lines comfortably
max_C_Model <- max(total_C_Model)

# Put into distinct df to 04_visualize.Rmd
## Always change according to current treatment
FB_FYM_total_C <- total_C_df
FB_FYM_final_pools <- final_pools_df
FB_FYM_rows_C_Model <- rows_C_Model
FB_FYM_rows_C_Spinup <- rows_C_Spinup
```

```{r plot, echo=FALSE}
# Plot individual pools and total SOC for the spin-up period
matplot(years_Model, Ct_Soil, type = "l", lty = 1, col = 1:5,
        xlab = "Time (years)", ylab = "C stocks (Mg/ha)",
        main = "Pools and Total SOC (Prediction)", ylim = c(0, max_C_Model * 1.1))
lines(years_Model, rows_C_Model, col = "black", lty = 2)
legend("topright", c("Total SOC", "DPM", "RPM", "BIO", "HUM", "IOM"),
       lty = c(1, 1, 1, 1, 1, 1), col = c(1:5, "black"), bty = "n")

# Save the plots
# ggsave("plots/farako_ba/treatments/pools_model_C-FB.png", width = 10, height = 6)
```

```{r write data}
# Save results as csv
## Watch out! overwrite = T
write.csv2(FB_FYM_rows_C_Model, file = "output/farako_ba/treatments/FB-FYM_rows_C_model.csv")
write.csv2(FB_FYM_final_pools, file = "output/farako_ba/treatments/FB-FYM_final_pools.csv")
write.csv2(FB_FYM_total_C, file = "output/farako_ba/treatments/FB-FYM_total_C.csv")
```
